Capturing the shape of our experiences and how we remember them

### 2018-08-20 paper/main.tex
saving progress

While our life experiences often change gradually, they can also change rapidly (e.g. walking through a doorway) \citep{Radv12}. Prior research suggests that these sharp transitions (coined in the literature as ``event boundaries'') during an experience may serve to descretize our memories into ``events'' \citep{Radv12, RadvZack11, BrunEtal18, HeusEtal18, ClewDava18}. Short timescale autocorrelations in the contents of an experience are thought associatively link information in memory, while more drastic changes in perceptual or semantic aspects of an experience may lead to discontinuities in memory \citep{HeusEtal18, BrunEtal18, Radv12}. There is now compelling evidence that event boundaries can influence memory organization when learning sequences of items \citep{HeusEtal18, DuBrDava13}, during navigation \citep{BrunEtal18}, and narrative comprehension \citep{ZwaaRadv98, EzzyDava11}.  However, whether people naturally recount their experiences in a way that is structured around events is not known.

@@

\subsubsection{The shape of a naturalistic experience is preserved in recall}
Classic approaches to studying free recall in episodic memory (such as overall accuracy and serial position curve) focus on the quantity of information recalled \citep{Murd62a}. More recent metrics (such as lag-CRP, semantic and temporal clustering) go further to describe list-level memory organization \citep{Kaha96, }. However, none of these approaches capture the rich autocorrelated structure inherent in naturalistic stimuli and associated recall of those experiences. In the next set of analyses, we test the hypothesis that successful recall of a naturalistic stimulus entails recovering the trajectory of a stimulus.

@@

To visualize the relationship between the video and recall event models, we embedded them into a 2D space (using the UMAP dimensionality reduction algorithm, for details see Methods) where the points represent video/recall events and the distance between them represents their similarity/distance in ``topic space'' (Fig.~\ref{fig:trajectory}a).  We observed that visually, it appears that the two models have a very similar temporal evolution and geometric structure. To further quantify this correspondence and to characterize how participants navigated through the space during recall, we analyzed the angle between successive recall events. We created a grid of evenly spaced points (.25 units) in the 2D ``topic space''. For each point on the grid, we drew a circle (radius=.25) around the point and grouped together all recall transitions that intersected the circle (across subjects). To visualize the average angle, we converted each transition angle to a unit vector and then averaged the vectors together. To assess consistency in the direction of the recall transition across participants, we performed a Rayleigh test (p$<$.001, corrected at p$<$.05 using permutation procedure described in Methods). Thus participants' recalls followed the same path as the video model (Fig.~\ref{fig:trajectory}b), suggesting the shape of the stimulus was preserved despite idiosyncratic differences between participants in their recalls.

@@

Just like in a traditional ``free-recall'' list-learning experiment where participants view a list of words and then verbally recall them, our video-recall matching analysis approach affords us the ability to analyze memory in the same way. We considered 6 classic list-learning analyses: overall accuracy, probability of first recall \citep{Hoga75, HowaKaha99; Lami99}, lag-conditional response probability \citep{Kaha96, HowaKaha99}, serial position curve \citep{Murd62a}, temporal clustering and semantic clustering \citep{HowaKaha02, PolyEtal09}. Extending these classic list-learning analyses to naturalistic stimuli allows to directly compare trial- and naturalistic-based recall studies (see Supplemental Materials Section 2 for details).

@@

\subsubsection{Choosing topic model parameters}
There were 3 critical parameters related to fitting the topic model: 1) the number of topics, 2) the window size of text descriptions of the video used to fit the model, and 3) the window size of recall sentences used to transform the recall data.  To chose these parameter values, we performed a grid search where the range of possible parameter values was 1, 5, 10, 25, 50, 100, 200, and 500. Our optimization objective was defined as the correlation between the hand annotated memory performance and the root mean squared distance between the video model and the recall model before any further processing (e.g. hidden Markov modelling, averaging within event, etc). While many of the parameter combination elicited moderately high correlations, the best choice was 100 topics, 50 video segments and 10 recall sentences.

\subsubsection{Extracting events using a hidden Markov model}
The topic model timepoint-by-timepoint correlation matrices all exhibited a block-diagonal structure (with small off-diagonal values), suggesting that the models were comprised of a number of sequential `states' (or events). To capture this structure, we fit the video and each recall model using a hidden Markov model (HMM). Given a number of states or events ($k$), the HMM recovers a set of labels that segments consecutive timepoints into $k$ events (REF markov paper). To implement this analysis, we used the Brainiak toolbox \citep{BaldEtal17}.

@@

\subsubsection{Visualizing the video and recall event models}
To better understand the temporal structure of the video event model (34 events by 100 topics) and the recall event models (8-27 events by 100 topics), we used a technique called UMAP to reduce the ``topic-space'' from 100 dimensions down to 2 dimensions (REF). UMAP is a nonlinear dimensionality reduction technique which models the manifold of the data with a fuzzy topological structure, and then searches for a (2D) projection of the data that has the closest equivalent fuzzy topological structure. We concatenated (vertically stacked) all event models (video, average recall, and individual recall), and then fit and transformed all of the models at once. This assured that the models were projected into the same space.

\subsubsection{Vector field analysis}
To quantify (and visualize) the flow of recall from event to event, we performed a vector field analysis.  We tiled the 2D topic space (x, y: -6 to 6 by .25) with an evenly spaced grid. For each grid point, we drew a square around the point (height/width=.5). Then, we tested whether any line segments (formed by event recall transitions) passed through this area of the topic space.  For example, say that a participant transitioned from recalling event 2 to event 3. These 2 recall events correspond to 2 points in topic space, and connecting them forms a line segment. We collected all line segments that passed through a given section of topic space (collapsing across participants). To plot the average direction of the line segments (i.e. the arrows for each grid point in ~\ref{fig:trajectory}b), we converted each of them to unit vectors and then averaged. For grid points where the direction was consistent (across all participants contributing to that point), the length of the arrow approaches 1, whereas if the direction was random the length of the arrow approaches 0. Lastly, we converted each unit vector to an angle (in radians) by taking the inverse tangent of the x, y components of the vector. To test whether the distribution of angles was significantly non-uniform (i.e. displayed a preferred direction across participants), we performed a Rayleigh test on the angles (REF). Arrows where the Rayleigh test was significant are displayed in color while non-significant tests are displayed in gray with lower opacity. Note that it was not computationally tractible to perform this analysis in the original 100 dimensional space because the number of grid points grows factorially with the number of topic dimensions.

\subsubsection{Topic vector word clouds}
 We created word clouds to visualize the themes contained in the recall events. One component of the topic model comprises a words (XX) by topics (100) matrix (R), where the rows represent the weight of a given word in each topic.  To find words that were maximally associated with a particular event vector, we computed the dot product between R and v, which gave a 1 by \# of words vector where the values represent the `activation' of each of in the event. Then, we created word clouds by extracting the top $n$ words and plotting them where the size of the word is proportional to its `activation' in the event.

>>>>

While our life experiences often change gradually, they can also change rapidly (e.g. walking through a doorway) \citep{RadvZack17}. Prior research suggests that these sharp transitions (coined in the literature as ``event boundaries'') during an experience may serve to descretize our memories into ``events'' \citep{RadvZack17, BrunEtal18, HeusEtal18, ClewDava17}. Short timescale autocorrelations in the contents of an experience are thought associatively link information in memory, while more drastic changes in perceptual or semantic aspects of an experience may lead to discontinuities in memory \citep{HeusEtal18, BrunEtal18}. There is now compelling evidence that event boundaries can influence memory organization when learning sequences of items \citep{HeusEtal18, DuBrDava13}, during navigation \citep{BrunEtal18}, and narrative comprehension \citep{ZwaaRadv98, EzzyDava11}.  However, whether people naturally recount their experiences in a way that is structured around events is not known.

@@

\subsubsection{The shape of a naturalistic experience is preserved in recall}
Classic approaches to studying free recall in episodic memory (such as overall accuracy and serial position curve) focus on the quantity of information recalled \citep{Murd62a}. More recent metrics (such as lag-CRP, semantic and temporal clustering) go further to describe list-level memory organization \citep{Kaha96, PolyEtal09}. However, these approaches do not capture the rich temporal structure inherent in naturalistic stimuli and associated recall of those experiences. In the next set of analyses, we test the hypothesis that successful recall of a naturalistic stimulus entails recovering the trajectory of a stimulus.

@@

To visualize the relationship between the video and recall event models, we embedded them into a 2D space using the UMAP dimensionality reduction algorithm ~\citep{McInHeal18}, where the points represent video/recall events and the distance between them represents their similarity/distance in ``topic space'' (Fig.~\ref{fig:trajectory}a).  We observed that visually, it appears that the two models have a very similar temporal evolution and geometric structure. To further quantify this correspondence and to characterize how participants navigated through the space during recall, we analyzed the angle between successive recall events. We created a grid of evenly spaced points (.25 units) in the 2D ``topic space''. For each point on the grid, we drew a circle (radius=.25) around the point and grouped together all recall transitions that intersected the circle (across subjects). To visualize the average angle, we converted each transition angle to a unit vector and then averaged the vectors together. To assess consistency in the direction of the recall transition across participants, we performed a Rayleigh test (p$<$.001, corrected at p$<$.05 using permutation procedure described in Methods). Thus participants' recalls followed the same path as the video model (Fig.~\ref{fig:trajectory}b), suggesting the shape of the stimulus was preserved despite idiosyncratic differences between participants in their recalls.

@@

Just like in a traditional ``free-recall'' list-learning experiment where participants view a list of words and then verbally recall them, our video-recall matching analysis approach affords us the ability to analyze memory in the same way. We considered 6 classic list-learning analyses: overall accuracy, probability of first recall \citep{Hoga75, HowaKaha99, Lami99}, lag-conditional response probability \citep{Kaha96, HowaKaha99}, serial position curve \citep{Murd62a}, temporal clustering and semantic clustering \citep{HowaKaha02, PolyEtal09}. Extending these classic list-learning analyses to naturalistic stimuli allows to directly compare trial- and naturalistic-based recall studies (see Supplemental Materials Section 2 for details).

@@

\subsubsection{Choosing topic model parameters}
There were 3 critical parameters related to fitting the topic model: 1) the number of topics, 2) the window size of text descriptions of the video used to fit the model, and 3) the window size of recall sentences used to transform the recall data.  To chose these parameter values, we performed a grid search where the range of possible parameter values was 1, 5, 10, 25, 50, 100, 200, and 500. Our optimization objective was defined as the correlation between the hand annotated memory performance and the root mean squared distance between the video model and the recall model before any further processing (e.g. hidden Markov modelling, averaging within event, etc). While many of the parameter combination elicited moderately high correlations, the optimal choice was 100 topics, 50 video segments and 10 recall sentences.

\subsubsection{Extracting events using a hidden Markov model}
The topic model timepoint-by-timepoint correlation matrices all exhibited a block-diagonal structure (with small off-diagonal values), suggesting that the models were comprised of a number of sequential `states' (or events). To capture this structure, we fit the video and each recall model using a hidden Markov model (HMM). Given a number of states or events ($k$), the HMM recovers a set of labels that segments consecutive timepoints into $k$ events ~\citep{Rabi89, BaldEtal17}. To implement this analysis, we used the Brainiak toolbox \citep{BaldEtal17}.

@@

\subsubsection{Visualizing the video and recall event models}
To better understand the temporal structure of the video event model (34 events by 100 topics) and the recall event models (8-27 events by 100 topics), we used a technique called UMAP ~\citep{McInHeal18} to reduce the ``topic-space'' from 100 dimensions down to 2 dimensions. UMAP is a nonlinear dimensionality reduction technique which models the manifold of the data with a fuzzy topological structure, and then searches for a (2D) projection of the data that has the closest equivalent fuzzy topological structure. We concatenated (vertically stacked) all event models (video, average recall, and individual recall), and then fit and transformed all of the models at once. This assured that the models were projected into the same space.

\subsubsection{Vector field analysis}
To quantify (and visualize) the flow of recall from event to event, we performed a vector field analysis.  We tiled the 2D topic space (x, y: -6 to 6 by .25) with an evenly spaced grid. For each grid point, we drew a square around the point (height/width=.5). Then, we tested whether any line segments (formed by event recall transitions) passed through this area of the topic space.  For example, say that a participant transitioned from recalling event 2 to event 3. These 2 recall events correspond to 2 points in topic space, and connecting them forms a line segment. We collected all line segments that passed through a given section of topic space (collapsing across participants). To plot the average direction of the line segments (i.e. the arrows for each grid point in ~\ref{fig:trajectory}b), we converted each of them to unit vectors and then averaged. For grid points where the direction was consistent (across all participants contributing to that point), the length of the arrow approaches 1, whereas if the direction was random the length of the arrow approaches 0. Lastly, we converted each unit vector to an angle (in radians) by taking the inverse tangent of the x, y components of the vector. To test whether the distribution of angles was significantly non-uniform (i.e. displayed a preferred direction across participants), we performed a Rayleigh test on the angles. Arrows where the Rayleigh test was significant are displayed in color while non-significant tests are displayed in gray with lower opacity.

\subsubsection{Topic vector word clouds}
 We created word clouds to visualize the themes contained in the recall events. One component of the topic model comprises a words (2117) by topics (100) matrix (R), where the rows represent the weight of a given word in each topic.  To find words that were maximally associated with a particular event vector, we computed the dot product between R and v, which gave a 1 by \# of words vector where the values represent the `activation' of each of in the event. Then, we created word clouds by extracting the top $n$ words and plotting them where the size of the word is proportional to its `activation' in the event.
